---
title: "Chapter 3 Discrete Random Variables"
output:
  word_document: default
  html_document: default
---


## Functions of random variables

There are many reasons why we might be more interested in looking at the distribution of a **function** of a random variable $X$ than the actual variable $X$. 

The probability mass function of $g(X)$ can be computed as follows. For each $y$, the probability that $g(X)=y$ is given by $\sum p(x)$ where the sum is over all values of $x$ such that $g(x)=y$.

#### Example
Suppose that $X$ is a binomial random variable with $n=4$ and $p=\frac{2}{3}$. What is the pmf of $Y=2X+1$. 

<https://media.csuchico.edu/media/Math+350+Binomial+Transformation/1_y4wihpgf>

```{r table1, echo=FALSE,message=FALSE,warning=FALSE}
tabl <- "

|-----|-----|-----|-----|-----|-----|
| X   |  0  | 1   | 2   | 3   |4    |
|-----|-----|-----|-----|-----|-----|
| Y   |  1  | 3   |   5 | 7   |  9  | 
|-----|-----|-----|-----|-----|-----|
|P(x) |.012 |.099 | .296|.395 |.198 |
|-----|-----|-----|-----|-----|-----|
"
cat(tabl) 

x<-0:4
n<-4
dbinom(x,4,2/3)
```

#### Theorem 3.7
Let $X$ be a discrete random variable with probability mass function $p$, and let $g$ be a function. Then 
\[
E[g(X)]=\sum g(x)p(x)
\]

#### Example
Suppose $X$ is a binomial random variable with $n=10$ and $p=\frac{2}{3}$. What is the expected value of $3X-4$?

<https://media.csuchico.edu/media/Math+350+Binomial+Expected+Value/1_9s7wna7g>

```{r}



```



### Theorem 3.8
For random variables $X$ and $Y$, and constants $a$, $b$, and $c$:
\[
E[aX+bY]=aE[X]+bE[Y]
\]
\[
E[c]=c
\]


#### Example
Repeat the above exercise using Theorem 3.8

$E(Y)=E(3X-4)=3E(X)-4=3*10*(2/3)-4=16$

#### End of Section Problems

1. Let $X$ be a discrete random variable with pmf given by 
P(X=0)=1/4, P(X=1)=1/2, P(X=2)=1/8, P(X=3)=1/8

a. Find the pmf of $Y=X-1$.

```{r}


```


b. Find the pmf of $U=X^{2}$.

```{r}


```


c. Find the pmf of $V=(X-1)^{2}$.

```{r}


```


2. Let $X$ and $Y$ be random variables such that $E[X]=2$ and $E[Y]=3$.

a. Find $E[4X+5Y]$.

```{r}


```


b. Find $E[4X-5Y+2]$.

```{r}


```


3. Marksmanship competition at a certain level requires each contestant to take ten shots with each of two different handguns. Final scores are computed by taking a weighted average of 4 times the number of bull-eyes made with the first gun plus 6 times the number gotten with the second gun. If Bertha has a 30% chance of hitting the bull's-eye with each shot from the first gun and a 40% chance with each shot from the second gun, what is her expected score?

```{r}



```




