---
title: "Chapter 3 Discrete Random Variables"
output:
  html_document: default
  word_document: null
---

```{r,message=FALSE,warning=FALSE}
library(dplyr)
```


## Section 3.1 Probability Mass Functions

Given a random experiment with an outcome space of $S$, a function $X$ that assigns one and only one real number $X(s)=x$ to each
element in $S$ is called a **random variable**.

#### Example  
Let the random experiment be the cast of a die.  Then the outcome space associated with this experiment is $S=\{1,2,3,4,5,6\}$, with the elements of $S$ indicating the number of spots on the side facing up.  The space of the random variable $X$ is then $\{1,2,3,4,5,6\}$.  If we associate the probability 1/6 with each outcome, then, Pr($X=5$)=1/6 and Pr($2\leq X\leq 5$)=4/6.


### Distribution of a Random Variable:

When a probability distribution has been specified on the sample space of an experiment, we can determine a probability distribution for the possible values of each random variable $X$.

This section is focused on probability distributions for discrete distributions.  It is said that $X$ has a discrete distribution if $X$ can only take the values of a finite number $k$ different numbers $x_{1},x_{2},\dots,x_{k}$ or at most, an infinite sequence of different values $x_{1},x_{2},\dots$.  Random variables that can take any value in an interval are called continuous and will be discussed in a later chapter. Working with discrete random variables requires **summation** while continuous random variables required **integration**.


Discrete variables are integers and usually represent a count of something while continuous variables take values in an interval of real numbers and often measure something.

## Section 3.1 Probability Mass Functions

### Definition 3.3
A discrete random variableis a variable that takes integer values and is characterized by a *probability mass function (pmf)*. The pmf *p* of a random variable $X$ is given by
\[
p(x)=P(X=x)
\]
The above equation can be read as: the probability that the random variable $X$ is equal to some value, $x$. 

What two properties does the pmf satisfy?:

1.

2.

To check that a function is a pmf, we check that all of its values are probabilities, and that those values sum to one.

After reading through  the Example 3.4 in the book, consider the following example:

#### Example

The great English diarist Samuel Pepys asked his friend Sir Isaac Newton the following questions:  Is it more likely to get at least one 6 when six dice are rolled, at least two 6's when twelve dice are rolled, or at least three 6's when eighteen dice are rolled?  Compute the three probabilities.

<https://media.csuchico.edu/media/Math+350+Section+3.1+3+Dice+Roll/1_01qrfb18>

Now do above problem using simulation in R:

```{r}
sixes_6<-replicate(10000,{
  roll<-sample(1:6,6,replace=TRUE)
  number_sixes<-roll==6
  at_least_1<-sum(number_sixes)>0
})

mean(sixes_6)

sixes_12<-replicate(10000,{
  roll<-sample(1:6,12,replace=TRUE)
  number_sixes<-roll==6
  at_least_1<-sum(number_sixes)>1
})

mean(sixes_12)

sixes_18<-replicate(10000,{
  roll<-sample(1:6,18,replace=TRUE)
  number_sixes<-roll==6
  at_least_1<-sum(number_sixes)>2
})

mean(sixes_18)


```

#### You try it
Ace-six flats are a type of crooked dice where the cube is foreshortened in the one-six direction, the effect being that 1's and 6's are more likely to occur than any of the other four faces.  Let $p(s)$ denote the probability that the face showing is $s$.  For many ace-six flats, the ``cube'' is more asymmetric to the extent that $p(1)=p(6)=1/4$, while $p(2)=p(3)=p(4)=p(5)=1/8$.  Is this a valid discrete probability function?   Explain.



#### End of Section Problems

1. Let $X$ be a discrete random variable with probability mass function given by

```{r table1, echo=FALSE,message=FALSE,warning=FALSE}
tabl <- "
|-----|-----|-----|-----|-----|
| X   | 0   | 1   | 2   | 3   |
|-----|-----|-----|-----|-----|
|p(x) |1/4  | 1/2 |1/8  | 1/8 |
|-----|-----|-----|-----|-----|
"
cat(tabl) 
```

a. Verify that $p$ is a valid probability mass function.


b. Find $P(X\geq 2)$.

```{r}
1/8 +1/8
```


c. Find $P(X\geq 2|X\geq 1)$.
```{r}
#o.25/(0.25+0.5)

```


d. Find $P(X\geq 2 \cup X\geq 1)$.
```{r} 
#P(x>=1)=(1/2)+(1/8)+(1/8)0.75
```


2. Let $X$ be a discrete random variable with probability mass function given by

```{r table2, echo=FALSE,message=FALSE,warning=FALSE}
tabl <- "
|-----|-----|-----|-----|-----|
| X   | 0   | 1   | 2   | 3   |
|-----|-----|-----|-----|-----|
|p(x) |1/4  | 1/2 |1/8  | 1/8 |
|-----|-----|-----|-----|-----|
"
cat(tabl) 
```

a. Use `sample` to create a sample of size 10,000 from $X$ and *estimate* $P(X=1)$ from you sample. Is your result close to the actual value?

```{r}
x<- 0:3
prob_x<-c(0.25,0.5,0.125,0.125)  

results<-sample(x,10000,replace =TRUE,prob = prob_x)
datafor1<-results==1
mean(datafor1)

```


b. Use `table` on your sample from part (a) to *estimate* the pmf of $X$ from your sample. You result should be similar to the pmf given in the problem.

```{r}

table(results)/10000

```

3. Let $X$ be a discrete random variable with pmf given by

```{r table3, echo=FALSE,message=FALSE,warning=FALSE}
tabl <- "
|-----|-----|-----|-----|
| X   | 0   | 1   | 2   | 
|-----|-----|-----|-----|
|p(x) |c/4  | c/2 | c   | 
|-----|-----|-----|-----|
"
cat(tabl) 
```

Find the value of $C$ that makes $p$ a valid probability mass function.





4. Suppose a particle moves along the x-axis beggining at 0. It moves one integer to the left or right with equal probability. What is the pmf of its position after four steps. Use simulation to find the pmf. 

```{r}



```

5. A gambling book recommends the following winning strategy for the game of roulette: Bet \$1 on red. If red appears (which has probability of 18/38), then take the dollar profit and quit. If red does not appear and you lose this bet, make additional \$1 bets on red on each of hte next two spins of the wheel and then quit. Let $X$ denoteyour winings when you quit. 

a. Give the probability mass function for $X$ (ie. give each possible value for $X$ and the probability associated with $X$).

```{r table4, echo=FALSE,message=FALSE,warning=FALSE}
tabl <- "
|-----|-----|-----|-----|
| X   |     |     |     | 
|-----|-----|-----|-----|
|p(x) |     |     |     | 
|-----|-----|-----|-----|
"
cat(tabl) 
```

b. Calculate $P(X>0)$.




6. A man buys a racehorse for \$20,000 and enters two races. He plans to sell the horse afterward, hoping to make a profit. If the horse wins both races, its value will jump to \$100,000. If it wins one of the races, it will be worth \$50,000. If the horse loses both races he will only be able to sell the horse for \$10,000. The man believes there is a 20\% chance that the horse will win the first race and a 30\% chance that the horse will win the second race. Assuming the two races are independent events, find the pmf for $X$ where $X$ is equal to the *profit*.

```{r,echo=FALSE,message=FALSE,warning=FALSE}
tabl <- "
|-----|-----|------|-----|
| X   |     |      |     | 
|-----|-----|------|-----|
|p(x) |     |      |     | 
|-----|-----|------|-----|
"
cat(tabl) 
```


non numeric 1.
```{r}
  #pmf after four steps
r<-c(-1,1)
parti <-replicate(10000,{
  steps<-sample(r,4,replace=TRUE) #this has the steps potentially taken
  results<-sum(steps) # this tells us where this person ended up
  
  
})
table(parti)/10000





```


non numeric 2
```{r}



```